---
title: "Creating shape files"
author: "Lauren Ponisio, based on a [Data Carpentary Lesson](https://datacarpentry.org/)"
output:
  pdf_document: default
  html_document: default
---

```{r setup, echo=FALSE}
source("setup.R")
```

```{r load-libraries, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
library(terra)
library(ggplot2)
library(dplyr)
library(sf)
```

This lesson will review how to import spatial points stored in `.csv` (Comma
Separated Value) format into R as an `sf` spatial object. We will also
reproject data imported from an ESRI `shapefile` format, export the reprojected 
data as an ESRI `shapefile`, and plot raster and vector data as layers in the same plot.

# Objectives

- Import .csv files containing x,y coordinate locations into R as a data frame.
- Convert a data frame to a spatial object.
- Export a spatial object to a text file.


```{r load-data, echo=FALSE, results="hide"}
## Load data we have used in previous demos
## roads
lines_HARV <- st_read("data/NEON-DS-Site-Layout-Files/HARV/HARV_roads.shp")
## area of interest boundary
aoi_boundary_HARV <- st_read("data/NEON-DS-Site-Layout-Files/HARV/HarClip_UTMZ18.shp")
## new, us states
country_boundary_US <- st_read("data/NEON-DS-Site-Layout-Files/US-Boundary-Layers/US-Boundary-Dissolved-States.shp")
## data collection towers
point_HARV <- st_read("data/NEON-DS-Site-Layout-Files/HARV/HARVtower_UTM18N.shp")
```


## Spatial Data in Text Format

The `HARV_PlotLocations.csv` file contains `x, y` (point) locations for study
plot where NEON collects data on
[vegetation and other ecological metics](https://www.neonscience.org/data-collection/terrestrial-organismal-sampling).
We would like to:

- Create a map of these plot locations.
- Export the data in an ESRI `shapefile` format to share with our colleagues. This
  `shapefile` can be imported into most GIS software.
- Create a map showing vegetation height with plot locations layered on top.

Spatial data are sometimes stored in a text file format (`.txt` or `.csv`). If
the text file has an associated `x` and `y` location column, then we can
convert it into an `sf` spatial object. The `sf` object allows us to store both
the `x,y` values that represent the coordinate location of each point and the
associated attribute data - or columns describing each feature in the spatial
object.

We will continue using the `sf` and `terra` packages in this episode.

## Import .csv

To begin let's import a `.csv` file that contains plot coordinate `x, y`
locations at the NEON Harvard Forest Field Site (`HARV_PlotLocations.csv`) and
look at the structure of that new object:

```{r read-csv}
plot_locations_HARV <-
  read.csv("data/NEON-DS-Site-Layout-Files/HARV/HARV_PlotLocations.csv")

str(plot_locations_HARV)
```

We now have a data frame that contains 21 locations (rows) and 16 variables
(attributes). Note that all of our character data was imported into R as
character (text) data. Next, let's explore the dataframe to determine whether
it contains columns with coordinate values. If we are lucky, our `.csv` will
contain columns labeled:

- "X" and "Y" OR
- Latitude and Longitude OR
- easting and northing (UTM coordinates)

Let's check out the column names of our dataframe.

```{r find-coordinates}
names(plot_locations_HARV)
```

## Identify X,Y Location Columns

Our column names include several fields that might contain spatial information.
The `plot_locations_HARV$easting` and `plot_locations_HARV$northing` columns
contain coordinate values. We can confirm this by looking at the first six rows
of our data.

```{r check-out-coordinates}
head(plot_locations_HARV$easting)
head(plot_locations_HARV$northing)
```

We have coordinate values in our data frame. In order to convert our data frame
to an `sf` object, we also need to know the CRS associated with those
coordinate values.

There are several ways to figure out the CRS of spatial data in text format.

1. We can check the file metadata in hopes that the CRS was recorded in the
  data.
2. We can explore the file itself to see if CRS information is embedded in the
  file header or somewhere in the data columns.

Following the `easting` and `northing` columns, there is a `geodeticDa` and a
`utmZone` column. These appear to contain CRS information (`datum` and
`projection`). Let's view those next.

```{r view-CRS-info}
head(plot_locations_HARV$geodeticDa)
head(plot_locations_HARV$utmZone)
```

It is not typical to store CRS information in a column. But this particular
file contains CRS information this way. The `geodeticDa` and `utmZone` columns
contain the information that helps us determine the CRS:

- `geodeticDa`: WGS84  -- this is geodetic datum WGS84
- `utmZone`: 18

To create the `proj4` associated with UTM Zone 18 WGS84 we can look up the
projection on the
[Spatial Reference website](https://www.spatialreference.org/ref/epsg/wgs-84-utm-zone-18n/),
which contains a list of CRS formats for each projection. From here, we can
extract the
[proj4 string for UTM Zone 18N WGS84](https://www.spatialreference.org/ref/epsg/wgs-84-utm-zone-18n/proj4/).

However, if we have other data in the UTM Zone 18N projection, it's much easier
to use the `st_crs()` function to extract the CRS in `proj4` format from that
object and assign it to our new spatial object. We've seen this CRS before with
our Harvard Forest study site (`point_HARV`).

```{r explore-units}
st_crs(point_HARV)
```

The output above shows that the points vector layer is in UTM zone 18N. We can
thus use the CRS from that spatial object to convert our non-spatial dataframe
into an `sf` object.

Next, let's create a `crs` object that we can use to define the CRS of our `sf`
object when we create it.

```{r crs-object}
utm18nCRS <- st_crs(point_HARV)
utm18nCRS

class(utm18nCRS)
```

## .csv to sf object

Next, let's convert our dataframe into an `sf` object. To do this, we need to
specify:

1. The columns containing X (`easting`) and Y (`northing`) coordinate values
2. The CRS that the column coordinate represent (units are included in the CRS) - stored in our `utmCRS` object.

We will use the `st_as_sf()` function to perform the conversion.

```{r convert-csv-shapefile}
plot_locations_sp_HARV <- st_as_sf(plot_locations_HARV,
                                   coords = c("easting", "northing"),
                                   crs = utm18nCRS)
```

We should double check the CRS to make sure it is correct.

```{r}
st_crs(plot_locations_sp_HARV)
```

## Plot Spatial Object

We now have a spatial R object, we can plot our newly created spatial object.

```{r plot-data-points}
ggplot() +
  geom_sf(data = plot_locations_sp_HARV) +
  ggtitle("Map of Plot Locations")
```

## In class: Import \& Plot Additional Points

We want to add two phenology plots to our existing map of vegetation plot
locations.

Import the .csv: `HARV/HARV_2NewPhenPlots.csv` into R and do the following:

1. Find the X and Y coordinate locations. Which value is X and which value is
   Y?
2. These data were collected in a geographic coordinate system (WGS84). Convert
  the dataframe into an `sf` object. HINT: check the crs of our other spatial objects to see if you can extract that crs to use to assign to the new data.
3. Plot the new points with the plot location points from above. Be sure to add
  a legend. Use a different symbol for the 2 new points!

If you have extra time, feel free to add roads and other layers to your map!

## Answers

1)


```{r}
  # first read in data
plot_locations_HARV_2 <-
  read.csv("data/NEON-DS-Site-Layout-Files/HARV/HARV_2NewPhenPlots.csv")
str(plot_locations_HARV_2)

# 1. find X and Y coord locations, which value is X and which is Y?

# getting the names of the columns
names(plot_locations_HARV_2)

# lat would be the latitude so Y
head(plot_locations_HARV_2$decimalLat)

# lon longitude so X
head(plot_locations_HARV_2$decimalLon)


```

2)

```{r}
# 2. convert df in an sf object, check crs of other spatial objects
      # to see if you can extract that crs to use to assign to the new data.
st_crs(country_boundary_US)
country_boundary_US_CRS <- st_crs(country_boundary_US)
country_boundary_US_CRS

class(country_boundary_US_CRS)


plot_locations_sp_HARV_2 <- st_as_sf(plot_locations_HARV_2,
                                   coords = c("decimalLon", "decimalLat"),
                                   crs = country_boundary_US_CRS)
str(plot_locations_sp_HARV_2)

```

3) 

```{r plot-locations-harv-orange}

# 3. plot new points with plot location from above, be sure to add a legend, use different symbol
          # for the two new points
ggplot() +
  geom_sf(data = plot_locations_sp_HARV, color = "hotpink") +
  geom_sf(data = plot_locations_sp_HARV_2, shape=5, color="forestgreen") + 
  ggtitle("Map of Plot Locations")


```


## Export to an ESRI `shapefile`

We can write an R spatial object to an ESRI `shapefile` using the `st_write` function
in `sf`. To do this we need the following arguments:

- the name of the spatial object (`plot_locations_sp_HARV`)
- the directory where we want to save our ESRI `shapefile` (to use `current = getwd()`
  or you can specify a different path)
- the name of the new ESRI `shapefile` (`PlotLocations_HARV`)
- the driver which specifies the file format (ESRI Shapefile)

We can now export the spatial object as an ESRI `shapefile`.

```{r write-shapefile, warnings="hide", eval=FALSE}
# allows us to write out an arcgis compatible shape file from a dataframe

st_write(plot_locations_sp_HARV,
         "data/PlotLocations_HARV.shp", driver = "ESRI Shapefile", append=FALSE)
# append = False will delete it and rewrite it 
```



## Summary

- Know the projection (if any) of your point data prior to converting to a
  spatial object.
- Convert a data frame to an `sf` object using the `st_as_sf()` function.
- Export an `sf` object as text using the `st_write()` function.


