---
title: "Network creation and visualization"
author: "Lauren Ponisio, based on workshop by Katya Ognyanova"
output:
  html_document: default
  pdf_document: default
---

<!-- Lesson Overview -->
# Computational Topics
- Make a graph from scratch
- Load in an edge list and create a graph 
- Visualize networks. 


```{r load-libraries, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
library(igraph)
library(dplyr)
library(networkD3)
```

In this lesson we will make some simple networks (i.e., graphs) in order to get a feeling for the structure of these type of data. We will also set the attributes of graphs (which we can eventually use for plotting).


# Demo part 1: Plotting an existing network

Let's begin by making a beautiful interactive graph of the interactions between characters in Le Mis.

```{r load-data, echo=FALSE, results="hide"}

```

How is the network data conveyed? 

```{r plot-leMis}
forceNetwork(Links = MisLinks, Nodes = MisNodes,
             Source = "source", Target = "target",
             Value = "value", opacityNoHover = 0.9,
             NodeID = "name",
             Group = "group", opacity = 0.8, zoom=TRUE)
```

Exciting! 

# Demo part 2: Making networks from scratch

We can also create networks from scratch in a whole variety of ways. 'make_full_graph' creates a graph where all of the vertices are connected to each other. 

```{r full-graph-creation}

```

We can also make a ring. 

```{r ring-graph-creation}

```

Or specify which vertices are connected. The default is that the edge are directed. Foodwebs are an example of a directed network. 

```{r graph-creation}

```

But we can make them undirected as well. Plant-pollinator interactions are an example of an undirected network (both species levels benefit from the interaction). 

```{r graph-creation-undirected}

```

Instead of vertex number we can use names. For example we can make a simple foodweb.

```{r graph-creation-named}

```
graphs have edges and vertices, accessible with the function 'E()' and 'V()'. We can also count the number of edges and vertices, which would be useful if we had a larger, unknown network. 

```{r,network-char}

```

We can use use $ to create attributes and get attributes. 
```{r, set-attributes}

```
We can then access those attributes. 
```{r, get-attributes}

```

We can manipulate a graph as an adjacency matrix.

```{r, get.adjacency}

```

# Demo part 3: Create a graph from OBA data, visualize it
Let's load in our OBA data and make a network for all the bumble bees in Oregon. We will need to subset the data, then keep only the bee and plant species columns. 

```{r, load-oba}

```

The entire network would be difficult to visualize, so let's subset the data to just bumble bees. We need to extract only the columns that are relevant for the bee-plant interactions. The column "Associated.plant" has the floral species a bee was caught visiting, if it was caught on a flower.

```{r, subset-obs}

```

There are a lot of blank rows from associated taxa for individuals not caught on a flower. We cannot do anything with those so we will just drop them. 

```{r, subset-obs-2}

```

Ugg, there are a lot of weird name add ones that will cause inconsistencies when we make the network because any unique name will be considered a unique node. Next week we will learn how to clean these out using regular expressions. For these week we will just drop them...

```{r, grep}

```


We can explore the species in our network. These will be the vertices. 
```{r, explore}

```
We have a variety of options for converting of data (which is basically an edge list) into a graph. One is to sum up our interactions but bee-plant combinations to make an bipartite adjacency matrix. Then convert that matrix to a igraph object. 

```{r, make-adj-1}

```

```{r, make-adj-2}

```

To use the 'forceNetwork' function we did for Le Mis, we need to convert our graph into a dataframe. Luckily there is a function 'igraph_to_networkD3' to convert from graph objects to what the network3d package wants. That function also wants a each vertex assigned to a group. We have a variety of options on how to assign groups, one is to look for modules or compartments in the network. 

```{r, plot-oba}
## find modules

## assign vertices to modules


## convert to a network 3d object

## plot the bumble-plant network

```

We can also use a flow-like visualization plot. 

```{r, plot-oba-2}

sankeyNetwork(Links = g_bumbles_net3d$links,
             Nodes = g_bumbles_net3d$nodes,
            Source = "source", Target = "target",
            Value = "value",  NodeID = "name", 
            nodePadding = 0, height = 5000)
```

# In class: Visualization challange 1.

Choose a different genus of bees and repeat the steps above to make a network. 

# In class: Visualization challange 2. 
How could we use graph attributes to create informative network visualizations? Set some attributes and use them to plot your network.
Consider using static network plots instead of network3D, see this [turorial](https://yunranchen.github.io/intro-net-r/advanced-network-visualization.html#visualization-for-static-network) for examples.  


```{r, plot-oba-challange}

```


